---
title: "Predicting distributional ranges of rocky intertidal species using a multi-dataset approach"
format: html
embed-resources: true
editor: visual
author: Simons D-L, Hipperson H, Webb TJ, Spencer M, Mieszkowska N
chunk_output_type: inline
toc: true
message: false
warning: false
---

## Introduction

This pipeline replicates all analyses in Simons *et al* (2025) "Predicting distributional ranges of rocky intertidal species using a multi-dataset approach".

Multiple resources have been used in this pipeline, including this [SDM guide](https://rspatial.org/sdm/index.html), Tom's Euro benthos habitat species guide, and `biomod2` tutorials.

## Set-up

### Load packages

```{r}
#| label: load-packages
#| echo: true
#| results: false

packages <- c("devtools",
              "qiime2R", #github download
              "microbiome", #github download
              "tidyverse",
              "vegan",
              "ggforce",
              "phyloseq",
              "cowplot",
              "geosphere",
              'lme4',
              "car",
              "sf",
              "pROC", #ROC and AUC for model validation 
              "ncdf4",
              "raster",
              "lubridate",
              "terra",
              "geodata",
              "predicts",
              "sdmpredictors",
              "tidyterra",
              "exactextractr",
              "arrow",
              "here",
              "emodnet.wfs", #github
              "biooracler",#github
              "biomod2",
              "dismo",
              "xgboost",
              "caret",
              "gam",
              "mda",
              "earth",
              "maxnet",
              "randomForest",
              "RODBC",
              "odbc") 

# Install any missing packages along with their dependencies
new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages, dependencies = TRUE)

# Load all packages
invisible(lapply(packages, library, character.only = TRUE))
```

### Functions

```{r}
#| label: get-nearest-coastline-function

get_nearest_coastline <- function(environment_stack, occurance_data, condition) {
  
  # convert to Spatraster
  env_terra <- rast(environment_stack)
  
  # Get CRS from raster
  raster_crs <- crs(env_terra, proj = TRUE)
  
  # Create original points as SpatVector
  points <- vect(occurance_data,
                 geom = c("lon", "lat"),
                 crs = raster_crs)
  
  # Extract environmental values at points
  presvals <- extract(env_terra, points)
  
  # Identify rows where all extracted values (except ID) are NA
  na_rows <- which(apply(presvals[, -1], 1, function(x) all(is.na(x))))
  
  # Create mask for coastline cells (where condition == 1)
  coastline_mask <- condition == 1
  
  # Extract points representing coastline cells
  coastline_points <- as.points(coastline_mask, values = FALSE)
  
  # Extract coordinates from original points
  coords <- crds(points)
  
  # For each NA point, find closest coastline point and replace coordinates
  for (i in na_rows) {
    this_point <- points[i]
    dists <- distance(this_point, coastline_points)
    nearest_idx <- which.min(dists)
    nearest_coords <- crds(coastline_points[nearest_idx])
    
    coords[i, ] <- nearest_coords
  }
  
  # Rebuild corrected SpatVector with updated coordinates
  corrected_points <- vect(data.frame(occurance_data, coords),
                           geom = c("x", "y"),
                           crs = crs(points))
  
  # Extract environmental variables at corrected points
  #presvals_corrected <- extract(env_terra, corrected_points)
  
  return(corrected_points)
}
```

```{r}
#| label: function-snap-to-valid-cell

snap_to_nearest_valid <- function(environment_stack, occurance_data) {
  
  # Select lon/lat columns using dplyr
  coords <- occurance_data %>% select(c(lon, lat))
  
  # Convert to SpatialPoints for raster extraction
  pts <- SpatialPoints(coords, proj4string = crs(environment_stack))
  
  # Extract raster values at occurrence points
  vals <- raster::extract(environment_stack, pts)
  
  # Identify points on any NA across layers
  na_points <- apply(vals, 1, function(x) any(is.na(x)))
  
  # Identify valid cells (no NAs in any layer)
  valid_cells <- Which(
    !is.na(environment_stack[[1]]) & !is.na(environment_stack[[2]]) & !is.na(environment_stack[[3]]),
    cells = TRUE
  )
  valid_xy <- xyFromCell(environment_stack, valid_cells)
  
  # Initialize corrected coordinates
  corrected_coords <- coords
  
  # Snap only points on NA cells
  if(any(na_points)) {
    nn <- FNN::get.knnx(valid_xy, coords[na_points, , drop = FALSE], k = 1)
    corrected_coords[na_points, ] <- valid_xy[nn$nn.index, ]
  }
  
  return(corrected_coords)
}
```

## Occurrence data

### Import paired eDNA and MarClim occurrences

Let's first read in our occurrences and metadata.

```{r}
#| label: import-combined-data

# occurance data
occurance_data <- read.csv(file = "Data/Occurrence_data/my_dataset/combined_methods_long_data.csv", row.names = 1)

# get lat long from meta data
metadata <- read.csv(file = "Data/Metadata/metadata.csv", na.strings = c(""))
metadata <- metadata %>%
  subset(select = c(localityID, exposure, decimalLongitude, decimalLatitude)) %>% unique() %>% na.omit()

# add lat long to occurance
occurance_data <- left_join(occurance_data, metadata)

# filter to only species of interest
species_list <- c(
  "Semibalanus balanoides",
  "Phorcus lineatus",
  "Chthamalus montagui",
  "Perforatus perforatus",
  "Patella depressa",
  "Bifurcaria bifurcata",
  "Asterias rubens"
)

occurance_data_reduced <- occurance_data %>% filter(taxa %in% species_list)
str(occurance_data_reduced)
```

Split into species of interest for combined and individual methods.

```{r}
#| label: filter-species-of-interest

# Clean species names to use in object names (e.g., "Semibalanus balanoides" -> "semibalanus")
clean_name <- function(name) {
  tolower(gsub(" ", "_", name))
}

# Loop through each species and create the filtered data frames
for (species in species_list) {
  clean <- clean_name(species)
  
  assign(paste0(clean, "_obs_both_methods"),
         occurance_data_reduced %>%
           filter(taxa == species, pa == 1) %>%
           select(taxa, decimalLongitude, decimalLatitude) %>%
           rename(lat = decimalLatitude, lon = decimalLongitude, species = taxa))

  assign(paste0(clean, "_obs_eDNA"),
         occurance_data_reduced %>%
           filter(taxa == species, method == "eDNA", pa == 1) %>%
           select(taxa, decimalLongitude, decimalLatitude) %>%
           rename(lat = decimalLatitude, lon = decimalLongitude, species = taxa))

  assign(paste0(clean, "_obs_visual"),
         occurance_data_reduced %>%
           filter(taxa == species, method == "Visual", pa == 1) %>%
           select(taxa, decimalLongitude, decimalLatitude) %>%
           rename(lat = decimalLatitude, lon = decimalLongitude, species = taxa))
}

# Check the data to make sure it loaded correctly
head(semibalanus_balanoides_obs_eDNA)
```

```{r}
#| label: define-latlong-extent

# Determine geographic extent of our data
# find general latitudinal and longitudinal boundaries

max_lat <- ceiling(max(occurance_data_reduced$decimalLatitude))
min_lat <- floor(min(occurance_data_reduced$decimalLatitude))
max_lon <- ceiling(max(occurance_data_reduced$decimalLongitude))
min_lon <- floor(min(occurance_data_reduced$decimalLongitude))

# Store boundaries in a single extent object
geographic_extent <- ext(x = c(min_lon, max_lon, min_lat, max_lat))
uk_extent <- ext(-11, 2, 49.5, 61)  # Westernmost to easternmost, southernmost to northernmost
uk_extent_match <- ext(-10.8,2,49.4,60.8)

# plot features
wrld <- world(path=".")
```

### Import wider MarClim occurrences

MarClim has a broader range of sites than just the sites matched to our eDNA samples. We can use this as an evaluation data set because it contains close to true absence and presence for the sampled sites. We are going to stick to 2022 - 2023 period to keep things simple.

```{r}
#| label: import-broad-marclim-data

# read
marclim_broad <- read.csv(file = "Data/Occurrence_data/my_dataset/MarClim_SACFOR_22_23.csv")

# filter for just semibalanus and chthamalus
marclim_broad_sp <- marclim_broad %>%
  select(Semibalanus.balanoides, 
         Chthamalus.montagui, 
         Lat..WGS84., 
         Long..WGS84.)

#convert to PA
marclim_broad_sp_PA <- marclim_broad_sp %>%
  mutate(across(
    c(Semibalanus.balanoides, Chthamalus.montagui),
    ~ ifelse(. %in% c("NR", "NS", "?", ""), 0, 1)
  )) %>% na.omit()

# tidy labels
marclim_broad_sp_PA <- marclim_broad_sp_PA %>%
  rename(
    lat = Lat..WGS84.,
    lon = Long..WGS84.
  )

#only within uK extent
marclim_broad_sp_PA <- subset(marclim_broad_sp_PA,
  lon >= -10.8 & lon <= 2 &
  lat >= 49.4 & lat <= 60.8
)

# make individual species presence only datasets for maps
semibalanus_marclim_broad <- marclim_broad_sp_PA %>%
  select(-c(Chthamalus.montagui)) %>% 
  filter(Semibalanus.balanoides == 1) %>% 
  mutate(species = "Semibalanus balanoides") %>% 
  select(-c(Semibalanus.balanoides))

chthamalus_marclim_broad <- marclim_broad_sp_PA %>%
  select(-c(Semibalanus.balanoides)) %>% 
  filter(Chthamalus.montagui == 1) %>% 
  mutate(species = "Chthamalus montagui") %>% 
  select(-c(Chthamalus.montagui))
```

### Import GBIF occurrences

We can also other occurrences from GBIF. These have been \# for now to save time and memory. They have been loaded directly in.

```{r}
#| label: download-from-GBIF-semibalanus

minYear = 2022

# Semibalanus

#semibalanus_GBIF_obs <- geodata::sp_occurrence("semibalanus", "balanoides", geo=TRUE, ext = uk_extent)
#dups <- duplicated(semibalanus_GBIF_obs[, c('lon', 'lat')])
#semibalanus_GBIF_obs_nodups <- semibalanus_GBIF_obs[!dups, ]
#semibalanus_GBIF_obs_nodups_recent <- subset(semibalanus_GBIF_obs_nodups, year >= minYear)
#dim(semibalanus_GBIF_obs_nodups_recent)

#write.csv(semibalanus_GBIF_obs_nodups_recent, file = paste("Data/Occurrence_data/GBIF/semibalanus_",minYear,".csv", sep = ""))

semibalanus_GBIF_clean <- read.csv("Data/Occurrence_data/GBIF/semibalanus_2022.csv")

# we only want 2022 and 2023 to match survey data
semibalanus_GBIF_clean <- semibalanus_GBIF_clean %>% 
  filter(year %in% c(2022, 2023))

# Phorcus

#phorcus_GBIF_obs <- geodata::sp_occurrence("phorcus", "lineatus", geo=TRUE, ext = uk_extent)
#dups <- duplicated(phorcus_GBIF_obs[, c('lon', 'lat')])
#phorcus_GBIF_obs_nodups <- phorcus_GBIF_obs[!dups, ]
##phorcus_GBIF_obs_nodups_recent <- subset(phorcus_GBIF_obs_nodups, year >= minYear)

#write.csv(phorcus_GBIF_obs_nodups_recent, file = paste("Data/Occurrence_data/GBIF/phorcus_",minYear,".csv", sep = ""))

phorcus_GBIF_clean <- read.csv("Data/Occurrence_data/GBIF/phorcus_2020.csv")

# Chthamalus

#chthamalus_GBIF_obs <- geodata::sp_occurrence("chthamalus", "montagui", geo=TRUE, ext = uk_extent)
#dups <- duplicated(chthamalus_GBIF_obs[, c('lon', 'lat')])
#chthamalus_GBIF_obs_nodups <- chthamalus_GBIF_obs[!dups, ]
#chthamalus_GBIF_obs_nodups_recent <- subset(chthamalus_GBIF_obs_nodups, year >= minYear)

#write.csv(chthamalus_GBIF_obs_nodups_recent, file = paste("Data/Occurrence_data/GBIF/chthamalus_",minYear,".csv", sep = ""))

chthamalus_GBIF_clean <- read.csv("Data/Occurrence_data/GBIF/chthamalus_2022.csv")

# we only want 2022 and 2023 to match survey data
chthamalus_GBIF_clean <- chthamalus_GBIF_clean %>% 
  filter(year %in% c(2022, 2023))

# Perforatus perforatus

#perforatus_GBIF_obs <- geodata::sp_occurrence("perforatus", "perforatus", geo=TRUE, ext = uk_extent)
#dups <- duplicated(perforatus_GBIF_obs[, c('lon', 'lat')])
#perforatus_GBIF_obs_nodups <- perforatus_GBIF_obs[!dups, ]
#perforatus_GBIF_obs_nodups_recent <- subset(perforatus_GBIF_obs_nodups, year >= minYear)

#write.csv(perforatus_GBIF_obs_nodups_recent, file = paste("Data/Occurrence_data/GBIF/perforatus_",minYear,".csv", sep = ""))

perforatus_GBIF_clean <- read.csv("Data/Occurrence_data/GBIF/perforatus_2020.csv")

# Patella depressa

#patella_GBIF_obs <- geodata::sp_occurrence("patella", "depressa", geo=TRUE, ext = uk_extent)
#dups <- duplicated(patella_GBIF_obs[, c('lon', 'lat')])
#patella_GBIF_obs_nodups <- patella_GBIF_obs[!dups, ]
#patella_GBIF_obs_nodups_recent <- subset(patella_GBIF_obs_nodups, year >= minYear)

##write.csv(patella_GBIF_obs_nodups_recent, file = paste("Data/Occurrence_data/GBIF/patella_",minYear,".csv", sep = ""))

patella_GBIF_clean <- read.csv("Data/Occurrence_data/GBIF/patella_2020.csv")

# Bifurcaria bifurcata

#bifurcaria_GBIF_obs <- geodata::sp_occurrence("Bifurcaria", "bifurcata", geo=TRUE, ext = uk_extent)
#dups <- duplicated(bifurcaria_GBIF_obs[, c('lon', 'lat')])
#bifurcaria_GBIF_obs_nodups <- bifurcaria_GBIF_obs[!dups, ]
#bifurcaria_GBIF_obs_nodups_recent <- subset(bifurcaria_GBIF_obs_nodups, year >= minYear)

#write.csv(bifurcaria_GBIF_obs_nodups_recent, file = paste("Data/Occurrence_data/GBIF/bifurcaria_",minYear,".csv", sep = ""))

bifurcaria_GBIF_clean <- read.csv("Data/Occurrence_data/GBIF/bifurcaria_2020.csv")
```

### Create data sets for model

We need 1) MarClim alone (as the evaluation data set), 2) GBIF alone as the visual data set, and 3) GBIF + eDNA.

```{r}
#| label: combining-data-sources-semibalanus

# combined GBIF and Marclim
semibalanus_GBIF_clean_reduced <- semibalanus_GBIF_clean %>% 
  mutate(species = as.factor(species),
         source = "GBIF") %>%
  select(species, lat, lon, source)

semibalanus_marclim_broad$source <- "MarClim"
final_visual_semibalanus <- rbind(semibalanus_GBIF_clean_reduced, semibalanus_marclim_broad)

# combining visual + eDNA (all)
semibalanus_balanoides_obs_eDNA$source <- "eDNA"
final_combined_semibalanus <- rbind(final_visual_semibalanus, semibalanus_balanoides_obs_eDNA) # plused 26 observations

# GBIF + eDNA
semibalanus_GBIF_eDNA <- rbind(semibalanus_GBIF_clean_reduced, semibalanus_balanoides_obs_eDNA) # plused 26 observations

# marclim alone 
semibalanus_marclim_broad <- semibalanus_marclim_broad %>% 
  select(c(species, lat, lon))

# GBIF alone
semibalanus_GBIF_clean_reduced <- semibalanus_GBIF_clean_reduced %>% 
  select(c(species, lat, lon)) 
```

We now have a few data sets for each species.

```{r}
#| label: view-final-datasets

head(semibalanus_marclim_broad) # marclim alone
head(semibalanus_GBIF_clean_reduced) #GBIF alone
head(semibalanus_GBIF_eDNA) #GBIF + eDNA
head(final_visual_semibalanus) #MarClim + GBIF
head(final_combined_semibalanus) #all visual + eDNA
```

Same again for chthamalus.

```{r}
#| label: combining-data-sources-chthamalus

# combined GBIF and Marclim
chthamalus_GBIF_clean_reduced <- chthamalus_GBIF_clean %>% 
  mutate(species = as.factor(species),
         source = "GBIF") %>%
  select(species, lat, lon, source)

chthamalus_marclim_broad$source <- "MarClim"
final_visual_chthamalus <- rbind(chthamalus_GBIF_clean_reduced, chthamalus_marclim_broad)

# combining visual + eDNA
chthamalus_montagui_obs_eDNA$source <- "eDNA"
final_combined_chthamalus <- rbind(final_visual_chthamalus, chthamalus_montagui_obs_eDNA) 

# GBIF + eDNA
chthamalus_GBIF_eDNA <- rbind(chthamalus_GBIF_clean_reduced, chthamalus_montagui_obs_eDNA) # plused 26 observations

# marclim alone 
chthamalus_marclim_broad <- chthamalus_marclim_broad %>% 
  select(c(species, lat, lon))

# GBIF alone
chthamalus_GBIF_clean_reduced <- chthamalus_GBIF_clean_reduced %>% 
  select(c(species, lat, lon)) 

# check
head(chthamalus_marclim_broad)
head(chthamalus_GBIF_clean_reduced)
head(chthamalus_GBIF_eDNA)
head(final_visual_chthamalus)
head(final_combined_chthamalus)
```

### Visualize occurrences

Let's plot our occurrences on a map. Grey points are sources from GBIF, green points are sourced from our generated MarClim data, and blue points are sources from our generated eDNA data.

```{r}
#| label: plot-occurance-GBIF-data-semibalanus-ggplot

# colours
GBIF_colour = "#D6B11F"
MarClim_colour = "#D44953"
eDNA_colour= "#70CBFF"

# Base ggplot map
semibalanus_map <- ggplot() +
  geom_sf(data = wrld, fill = "grey98", color = "lightgray") +
  
  # Add all points with color mapped to 'source'
  geom_jitter(data = final_combined_semibalanus, 
              aes(x = lon, y = lat, color = source), 
              size = 1.1, width = 0.025, height = 0.025, alpha = 0.6) +
  
  coord_sf(xlim = c(-11, 3), ylim = c(49, 61)) +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude", color = "Data source") +
  scale_color_manual(values = c(
    "GBIF" = GBIF_colour,
    "MarClim" = MarClim_colour,
    "eDNA" = eDNA_colour
  )) +
  theme(legend.position = "none")

semibalanus_map

# Save the plot
ggsave(semibalanus_map, filename = "Figures/semibalanus_map.png", width = 8, height = 6, dpi = 300)
```

```{r}
#| label: plot-occurance-GBIF-data-chthamalus-ggplot

# Base ggplot map
chthamalus_map_legend <- ggplot() +
  geom_sf(data = wrld, fill = "grey98", color = "lightgray") +
  
  # need to jitter points a bit so that some are not hidden
  geom_jitter(data = final_combined_chthamalus, 
              aes(x = lon, y = lat, color = source), 
              size = 1.1, width = 0.025, height = 0.025, alpha = 0.6) +
  
  coord_sf(xlim = c(-11, 3), ylim = c(49, 61)) +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude", color = "Data source") +
  scale_color_manual(values = c(
    "GBIF" = GBIF_colour,
    "MarClim" = MarClim_colour,
    "eDNA" = eDNA_colour
  )) +
  theme(legend.position = "bottom")

chthamalus_map <- ggplot() +
  geom_sf(data = wrld, fill = "grey98", color = "lightgray") +
  
  # need to jitter points a bit so that some are not hidden
  geom_jitter(data = final_combined_chthamalus, 
              aes(x = lon, y = lat, color = source), 
              size = 1.1, width = 0.025, height = 0.025, alpha = 0.6) +
  
  coord_sf(xlim = c(-11, 3), ylim = c(49, 61)) +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude", color = "Data source") +
  scale_color_manual(values = c(
    "GBIF" = GBIF_colour,
    "MarClim" = MarClim_colour,
    "eDNA" = eDNA_colour
  )) +
  theme(legend.position = "none")

chthamalus_map

# Save the plot
ggsave(chthamalus_map, filename = "Figures/chthamalus_map.png", width = 8, height = 6, dpi = 300)
```

```{r}
#| label: get-legend-for-plot

g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
}

legend <- g_legend(chthamalus_map_legend)
```

```{r}
#| label: joint-map-twospecies

joint_map <- cowplot::plot_grid(
  cowplot::plot_grid(semibalanus_map, chthamalus_map, ncol = 2, labels = c("a", "b"), rel_heights = c(0.8, 1)), legend, ncol =1, rel_heights = c(1, 0.06))


joint_map

# Save the plot
ggsave(joint_map, filename = "Figures/chthamalus_semibalanus_map.png", width = 8, height = 6, dpi = 300)
```

```{r}
#| label: joint-map-twospecies-long

joint_map_long <- cowplot::plot_grid(semibalanus_map, chthamalus_map, labels = c("a", "b"),rel_widths = c(1,1.5), ncol = 1, align = "v" )

joint_map_long

# Save the plot
ggsave(joint_map_long, filename = "Figures/chthamalus_semibalanus_map_long.png", width = 6, height = 11, dpi = 300)
```

```{r}
#| label: plot-occurance-GBIF-data-phorcus

# Set the file path and open the PNG device
png(filename = paste("Figures/phorcus_occs_map_", minYear, ".png", sep = ""),
    width = 800, height = 600)

# Plot commands
plot(wrld, xlim = c(-11, 3), ylim = c(49, 61), col = "light yellow", border = "light gray")

# Add GBIF points
points(phorcus_GBIF_clean$lon, 
       phorcus_GBIF_clean$lat, 
       col = 'darkgrey', 
       pch = 20)

# Add your points
points(x = phorcus_lineatus_obs_both_methods$lon, 
       y = phorcus_lineatus_obs_both_methods$lat, 
       col = "red", 
       pch = 20, 
       cex = 0.75)

# Close the device
dev.off()
```

```{r}
#| label: plot-occurance-GBIF-data-perforatus

# Set the file path and open the PNG device
png(filename = paste("Figures/perforatus_occs_map_", minYear, ".png", sep = ""),
    width = 800, height = 600)

# Plot commands
plot(wrld, xlim = c(-11, 3), ylim = c(49, 61), col = "light yellow", border = "light gray")

# Add GBIF points
points(perforatus_GBIF_clean$lon, 
       perforatus_GBIF_clean$lat, 
       col = 'darkgrey', 
       pch = 20)

# add my points
points(x = perforatus_perforatus_obs_both_methods$lon, 
       y = perforatus_perforatus_obs_both_methods$lat, 
       col = "red", 
       pch = 20, 
       cex = 0.75)

# Close the device
dev.off()
```

```{r}
#| label: plot-occurance-GBIF-data-patella

# Set the file path and open the PNG device
png(filename = paste("Figures/patella_occs_map_", minYear, ".png", sep = ""),
    width = 800, height = 600)

# Plot commands
plot(wrld, xlim = c(-11, 3), ylim = c(49, 61), col = "light yellow", border = "light gray")

# Add GBIF points
points(patella_GBIF_clean$lon, 
       patella_GBIF_clean$lat, 
       col = 'darkgrey', 
       pch = 20)

# add my points
points(x = patella_depressa_obs_both_methods$lon, 
       y = patella_depressa_obs_both_methods$lat, 
       col = "red", 
       pch = 20, 
       cex = 0.75)

# Close the device
dev.off()
```

```{r}
#| label: plot-occurance-GBIF-data-Bifurcaria

# Set the file path and open the PNG device
png(filename = paste("Figures/bifurcaria_occs_map_", minYear, ".png", sep = ""),
    width = 800, height = 600)

# Plot commands
plot(wrld, xlim = c(-11, 3), ylim = c(49, 61), col = "light yellow", border = "light gray")

# Add GBIF points
points(bifurcaria_GBIF_clean$lon, 
       bifurcaria_GBIF_clean$lat, 
       col = 'darkgrey', 
       pch = 20)

# add my points
points(x = bifurcaria_bifurcata_obs_both_methods$lon, 
       y = bifurcaria_bifurcata_obs_both_methods$lat, 
       col = "red", 
       pch = 20, 
       cex = 0.75)

# Close the device
dev.off()
```

### Check for sampling bias

```{r}
#| label: sampling-bias-points

# make a SpatVector
semibalanus_spat <- vect(final_visual_semibalanus, geom=c("lon", "lat"), crs="+proj=longlat +datum=WGS84")
class(semibalanus_spat)

# create a SpatRaster with the extent of acgeo
r <- rast(semibalanus_spat)

# set the resolution of the cells to (for example) 1 degree
res(r) <- 1

# extend (expand) the extent of the SpatRaster a little
r <- extend(r, ext(r)+1)

# sample:
set.seed(13)
acsel <- spatSample(semibalanus_spat, size=1, "random", strata=r)

# to illustrate the method and show the result
p <- as.polygons(r)
plot(p, border='gray')
points(semibalanus_GBIF_clean)
points(acsel, cex=1, col='red', pch='x') # selected points in red
```

## Absence and presence points (background points)

Let's sample background points within a radius of presence points (as seen in VanDerWal et al, 2009).

***Grey points*** are a random sample of 1000 points taken from within 30km around presence points. One point is sampled per raster cell to avoid spatial bias. These represent **candidate pseudo-absence locations** i.e. background points from within the accessible area.

***Blue points*** are a subset of the grey points. These are points whose raster cell center falls within the buffer circles. They represent **final valid** **pseudo-absences**.

```{r}
#| label: sample-psudo-absences

# circles with a radius of 30 km
x <- raster::buffer(semibalanus_spat, 30000)
pol <- aggregate(x)

# take a random sample of points within the polygons. 
# We only want one point per grid cell, sampling randomly from all circles
set.seed(999)
samp1 <- spatSample(pol, 1000, "random")

# no raster cells multiple times
pcells <- cells(r, samp1)
pcells <- unique(pcells[,2]) # remote duplicates
xy <- xyFromCell(r, pcells) # back to coordinates

# choose to select only those cells that have centers within circles
spxy <- vect(xy, crs="+proj=longlat +datum=WGS84")
xyInside <- intersect(spxy, x)

# plots to inspect results
plot(pol, axes=TRUE)
points(samp1, pch="+", cex=.5)
points(xyInside, cex=0.75, pch=20, col='blue')
```

## Environmental data

It's time to find environmental variables of interest. OCLE (<https://ocle.ihcantabria.com/>) provides meteo-oceanographic data for UK coastal communities, some of which are projected under future climate change scenarios.

To match the resolution of the projected estimates, we are going to download annual mean for each variable. Summer mean is available for historic periods, but not for projected periods.

```{r}
#| label: plot-features

# set uk extent and colours
uk_extent <- extent(-11, 2, 49.5, 61)
cols <- colorRampPalette(c("blue", "lightblue", "yellow", "red"))(100)
```

### Temp, pH, coastline (Bio-ORCLE)

Bio-ORACLE is a comprehensive data set providing a wide range of marine environmental variables tailored for ecological and biogeographical analysis. These variables include surface and benthic values for parameters such as temperature, salinity, nutrients, oxygen, pH, and more, derived from satellite and modeled oceanographic data.

We use the `biooracler` package here to obtain our environmental variables of interest from current and projected periods. We just load the datasets and skip running this code for now to save memory.

```{r}
#| label: variables-from-biooracle
#| eval: false
#| include: false

# explore layers
biooracler_layers <- biooracler::list_layers()

# define lat long
uk_lat <- c(49.39,60.8626)
uk_long <- c(-10.9968,2.5643)

# example of selecting five variables from three datasets
datasets <- list(

  #air temps
  list(dataset_id = "tas_baseline_2000_2020_depthsurf",
       variables = "tas_mean",
       constraints = 
         list(#time = c('2000-01-01T00:00:00Z', '2010-01-01T00:00:00Z'), 
              latitude = uk_lat, longitude = uk_long )),
  
  list(dataset_id = "tas_ssp585_2020_2100_depthsurf",
       variables = "tas_mean",
       constraints = 
         list(#time = c('2020-01-01T00:00:00Z', '2090-01-01T00:00:00Z'), 
              latitude = uk_lat, longitude = uk_long )),
  
  #ocean temps
  list(dataset_id = "thetao_baseline_2000_2019_depthsurf",
       variables = "thetao_mean",
       constraints = list(
         #time = c('2000-01-01T00:00:00Z', '2010-01-01T00:00:00Z'), 
         latitude = uk_lat, longitude = uk_long )), 
  
  list(dataset_id = "thetao_ssp585_2020_2100_depthsurf",
       variables = "thetao_mean",
       constraints = list(
         #time = c('2000-01-01T00:00:00Z', '2010-01-01T00:00:00Z'), 
         latitude = uk_lat, longitude = uk_long )), 
  
  # pH
  list(dataset_id = "ph_baseline_2000_2018_depthsurf",
       variables = "ph_mean",
       constraints = list(
         #time = c('2000-01-01T00:00:00Z', '2010-01-01T00:00:00Z'), 
         latitude = uk_lat, longitude = uk_long )),
  
  list(dataset_id = "ph_ssp585_2020_2100_depthsurf",
       variables = "ph_mean",
       constraints = list(
         #time = c('2000-01-01T00:00:00Z', '2010-01-01T00:00:00Z'), 
         latitude = uk_lat, longitude = uk_long )),
  
  list(dataset_id = "terrain_characteristics",
       variables = "coastline",
       constraints = list(
         #time = c('2000-01-01T00:00:00Z', '2010-01-01T00:00:00Z'), 
         latitude = uk_lat, longitude = uk_long ))
  
  )

# download layers
for (dataset in datasets) {
  
  dataset_id <- dataset$dataset_id
  variables <- dataset$variables
  constraints <- dataset$constraints
  
  layer <- download_layers(dataset_id, variables = variables, constraints = constraints)
  
  # Save using dataset_id
  save_path <- file.path("Data/Environmental_data/BIO-ORACLE", paste0(dataset_id, ".rds"))
  saveRDS(layer, save_path)
}
```

```{r}
#| label: read-and-explore-variables

## ocean temp
ocean_temp_present = readRDS("Data/Environmental_data/BIO-ORACLE/thetao_baseline_2000_2019_depthsurf.rds")
plot(
  ocean_temp_present,
  col = cols
)

ocean_temp_projected = readRDS("Data/Environmental_data/BIO-ORACLE/thetao_ssp585_2020_2100_depthsurf.rds")
plot(
  ocean_temp_projected,
  col = cols
)

## air temp
air_temp_present = readRDS("Data/Environmental_data/BIO-ORACLE/tas_baseline_2000_2020_depthsurf.rds")
plot(
  air_temp_present,
  col = cols
)

air_temp_projected = readRDS("Data/Environmental_data/BIO-ORACLE/tas_ssp585_2020_2100_depthsurf.rds")
plot(
  air_temp_projected,
  col = cols
)

## ph
ph_present = readRDS("Data/Environmental_data/BIO-ORACLE/ph_baseline_2000_2018_depthsurf.rds")
plot(ph_present,
     col = cols)

ph_projected = readRDS("Data/Environmental_data/BIO-ORACLE/ph_ssp585_2020_2100_depthsurf.rds")
plot(ph_projected,
     col = cols)

# terrain
terrain <- readRDS("Data/Environmental_data/BIO-ORACLE/terrain_characteristics.rds")
plot(terrain,
     main = "Terrain")
```

### Rocky substrate (EMODnet & JNCC)

We can use the package `emodnet.wfs` to obtain a full UK habitat map from EMODnet servers. Documentation for this service is [here](https://docs.ropensci.org/emodnet.wfs/).

```{r}
#| label: import-from-emodnet_wfs
#| eval: false
#| include: false

# service information 
#services <- emodnet_wfs()
#services[, c("emodnet_thematic_lot", "service_name")]

# initialising a client is the first step to interacting with an EMODnet Web Feature Services
#wfs_geo <- emodnet_init_wfs_client(service = "geology_seabed_substrate_maps")
#wfs_seabed <- emodnet_init_wfs_client(service =  "seabed_habitats_general_datasets_and_products")

# List contents of a WFS
#layer_info <- emodnet.wfs::emodnet_get_wfs_info(wfs_seabed)
#layer_info

# get descripion
#layer_info %>%
  #filter(layer_name == "ukseamap_latest_habitats") %>%
  #pull(abstract) %>%
  #cat()

# only UKASH Mosaic of Localised Maps includes intertidal zone

# define layer
#layers <- c("ukseamap_latest_habitats") 

# get the data
#emodnet_substrate <- emodnet_get_layers(wfs = wfs_seabed,
                                        #layers = layers,
                                        #simplify = TRUE)
# explore the data
#emodnet_substrate
#unique(colnames(emodnet_substrate))
#unique(emodnet_substrate$eunis_name)
#unique(emodnet_substrate$country)

# get hard rock only - no littoral :(
#rock_hard_substrate <- emodnet_substrate %>%
  #filter(str_detect(str_to_lower(eunis_name), "rock"))

#plot (takes horrendously long)
#ggplot(rock_hard_substrate) +
  #geom_sf(aes(fill = eunis_name)) +
  #scale_fill_viridis_d(option = "plasma", name = "Substrate Type") +
  #coord_sf(crs = 4326) +
  #theme_minimal() +
  #labs(
    #title = "EMODnet Habitat Map",
    #x = "Longitude", y = "Latitude"
  #)
```

Unfortunately, we couldn't get intertidal data from the above products. Let's try the JNCC UKASH Mosaic of Localized Maps which does include intertidal data, but only for 12% of the UK.

```{r}
#| label: read-in-UKASH-mosaic
#| eval: false
#| include: false

# Path to the unzipped .gdb folder for UKASH Mosaic of Localised Maps (with intertidal zone)
gdb_path <- "Data/Environmental_data/JNCC/UKASH_MosaicOfLocalisedMaps_v2025.gdb"

# List available layers
st_layers(gdb_path)

# read in
ukash <- st_read(gdb_path, layer = "UKASH_MosaicLocalisedMaps_v2025_1")

# explore
print(ukash)

# filter for rocky only
rock_ukash <- ukash %>%
  filter(str_detect(str_to_lower(OrigName), "rock"))

# make a raster and match format to the other environmental variables
rock_ukash_rast <- raster::rasterize(rock_ukash, ocean_temp_present)

# read in
rocky_substrate_mosaic

# plot
rock_ukash_plot <- ggplot(rock_ukash) +
  geom_sf(aes(fill = OrigName)) +
  scale_fill_viridis_d(option = "plasma", name = "Substrate Type") +
  coord_sf(crs = 4326) +
  theme_minimal() +
  labs(
    title = "UKASH Mosaic Localised Rock Habitat Map",
    x = "Longitude", y = "Latitude"
  ) + theme(legend.position = "none")

rock_ukash_plot

ggsave(rock_ukash_plot, filename = "Figures/UKASH_Mosaic_Localised_Rock.jpeg")
```

To get a better coverage, let's import rocky habitat information from JNCC Marine Recorder Public UK snapshot.

Two Microsoft access databases (primary and secondary) are available at <https://hub.jncc.gov.uk/assets/b9934e31-39b6-41f9-9364-d1e93db68307>. The rocky habitat we are interested in are defined here - <https://mhc.jncc.gov.uk/biotopes/jnccmncr00000003>. Primary database contains locations, secondary contains biotypes.

```{r}
#| label: read-in-JNCC-snapshot-data
#| eval: false
#| include: false

# lists the drivers already installed for excel, access ect...
unique(odbc::odbcListDrivers()[[1]])

# set database locations
primary_dbname <- "Data/Environmental_data/JNCC/SnapshotDatav52_PUBLIC_20220124.mdb"
secondary_dbname <- "Data/Environmental_data/JNCC//SnapshotDatav52_PUBLIC_20220124_secondary.mdb"

# Establish connections between R and Access
RODBC::odbcCloseAll() # clear any existing
primary_con <- RODBC::odbcConnectAccess2007(primary_dbname)
secondary_con <- RODBC::odbcConnectAccess2007(secondary_dbname)

RODBC::sqlTables(primary_con)
RODBC::sqlTables(secondary_con)

# Fetch the data from the SampleBiotopeAll table
locations <- RODBC::sqlFetch(primary_con, "Location", rownames = T)
SampleBiotopeAll <- RODBC::sqlFetch(secondary_con, "SampleBiotopeAll", rownames = T)

# Pull all the Biotope occurrence keys littoral habitats
Biotope_littoral <- SampleBiotopeAll %>%
  dplyr::filter(grepl("LR", BiotopeCode))  # matches any LR entries (includes eulittoral)

# join
biotype_location <- left_join(Biotope_littoral, locations, by = "ObjectId")
biotype_location_reduced <- biotype_location %>% select(ObjectId, BiotopeCode, Lat, Long, LatWGS84, LongWGS84, Qualifier, CoordinateSystem) %>% na.omit()

# remove uncertain 
biotype_location_reduced_certain <- biotype_location_reduced %>%
  dplyr::filter(!grepl("Uncertain", Qualifier))

#visualise - looks like a some points are incorrect
rocky_map <- ggplot() +
  geom_sf(data = wrld, fill = "lightyellow", color = "lightgray") +
  
  # Add all points with color mapped to 'source'
  geom_jitter(data = biotype_location_reduced, 
              aes(x = LongWGS84, y = LatWGS84, colour = Qualifier), 
              size = 0.8, width = 0.025, height = 0.025) +
  coord_sf(xlim = c(-11, 3), ylim = c(49, 61)) +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") #+
  theme(legend.position = "none")

rocky_map

ggsave(rocky_map, filename = "Figures/JNCC_snapshot_littoral_map.jpeg")
```

```{r}
#| label: remove-offshore-points-rocky
#| eval: false
#| include: false

# Convert raster coastline (value=1 for land) to polygons
coast_poly <- as.polygons(terrain, dissolve = TRUE, values = TRUE)

# Keep only the polygons where coastline == 1
coast_poly <- coast_poly[coast_poly$coastline == 1, ]

# Buffer (in degrees — ~0.01° ≈ 1.1 km at equator)
buffer_dist <- 0.03  # adjust based on desired intertidal width
coast_poly_buffered <- terra::buffer(coast_poly, width = buffer_dist)

# Make a SpatVector from Lat/Long WGS84
biotype_vect <- vect(
  biotype_location_reduced,
  geom = c("LongWGS84", "LatWGS84"), 
  crs = "OGC:CRS84"
)

# Intersect points with buffered polygon
biotype_onshore_vect <- terra::intersect(biotype_vect, coast_poly_buffered)

# Convert points and polygon to sf
biotype_sf <- st_as_sf(biotype_onshore_vect)
coast_sf   <- st_as_sf(coast_poly_buffered)

# Plot
jncc_snapshot_littorial_tidy_map <- ggplot() +
  geom_sf(data = coast_sf, fill = "lightgray", color = "lightblue") +
  geom_sf(data = biotype_sf, size = 1, color = "orange") +
  coord_sf(xlim = c(-11, 3), ylim = c(49, 61)) + # adjust to your study area
  theme_minimal() +
  theme(legend.position = "none")

jncc_snapshot_littorial_tidy_map

ggsave(jncc_snapshot_littorial_tidy_map, filename = "Figures/jncc_snapshot_littorial_tidy_map.jpeg", width = 5, height= 8)
```

Let's combined the two JNCC sources so we have the best cover for our environmental rocky variable.

```{r}
#| label: join-JNCC-rock-sources
#| eval: false
#| include: false

# Ensure CRS matches
biotype_sf <- st_transform(biotype_sf, crs(rock_ukash_rast))

# Keep biotype_sf for later use (this is your vector dataset)

# Convert sf to SpatVector for rasterising
biotype_vect <- vect(biotype_sf)

# Create a template raster
rocky_template <- rast(rock_ukash_rast)

# Rasterise points
rocky_rast <- rasterize(biotype_vect, rocky_template, field = 1, background = NA)

# Combine with existing rock raster
combined_rock_rast <- cover(rocky_rast, rock_ukash_rast)

# Plot
plot(combined_rock_rast, main = "Rocky substrate (binary)", col = "grey10")
res(combined_rock_rast) #0.05 x 0.05

# expand radius
# Calculate distance from rocky areas (in map units, e.g., meters)
dist_rast <- distance(combined_rock_rast)

# Keep only cells within 5 km
expanded_rock <- dist_rast <= 8000

# Convert to binary (1 = rocky/within 1 km, NA = non-rocky)
expanded_rock[expanded_rock == 0] <- NA
expanded_rock[expanded_rock == 1] <- 1

plot(expanded_rock, main = "Expanded rocky substrate", col = "grey10")

# Convert combined raster back to sf polygons if you want an sf version
#combined_rock_sf <- st_as_sf(as.polygons(combined_rock_rast, dissolve = TRUE, values = TRUE))

# keep 1 as 1, set everything else to NA
#rock01 <- ifel(combined_rock_rast == 1, 1, NA)

# polygonize only the 1s
#pol1 <- as.polygons(rock01, dissolve = TRUE, na.rm = TRUE)
#combined_rock_sf <- st_as_sf(pol1)

# optional cleanup
#combined_rock_sf <- sf::st_make_valid(combined_rock_sf)
#combined_rock_sf <- sf::st_cast(combined_rock_sf, "MULTIPOLYGON")

# make a raster and match format to the other environmental variables
#rock_rast_combined <- raster::rasterize(combined_rock_sf, ocean_temp_present
```

### Wave fetch (Burrows)

Wave fetch data is downloaded from <https://www.sams.ac.uk/people/researchers/burrows-professor-michael/>.

```{r}
#| label: wave-fetch

# Import
fetch <- rast("Data/Environmental_data/wave_fetch.asc")

# Adjust 'fact' depending on raster size (larger = faster, less detail)
fetch_small <- aggregate(fetch, fact = 10, fun = mean)

#Set land cells (value = 1) and open ocean (value = 0) to NA
fetch_small[fetch_small == 1] <- NA
fetch_small[fetch_small == 0] <- NA

# Convert downsampled raster to data frame
fetch_df <- as.data.frame(fetch_small, xy = TRUE)

# Check column names
colnames(fetch_df)  # usually "x", "y", "layer"

# Plot with ggplot
fetch_plot <- ggplot(fetch_df, aes(x = x, y = y, fill = wave_fetch)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_equal() +
  theme_minimal() +
  labs(fill = "Wave Fetch",
       title = "Wave Fetch Raster")

fetch_plot

# make same as environmental data
# Convert combined raster back to sf polygons if you want an sf version
fetch_sf <- as.polygons(fetch_small, dissolve = FALSE)  # dissolve=FALSE keeps each cell separate
fetch_sf <- st_as_sf(fetch_sf)   

#Check polygon CRS; if missing, assign the original projected CRS
st_crs(fetch_sf) <- 27700  # Example: British National Grid, adjust if different

# Widen buffer by 1000 m (example)
wf_buffered <- st_buffer(fetch_sf, dist = 1000)

#Reproject polygons to match raster CRS
fetch_sf_proj <- st_transform(wf_buffered, crs(ocean_temp_present))

# Rasterize
fetch_raster <- rasterize(vect(fetch_sf_proj), ocean_temp_present, field = "wave_fetch", background = NA)
```

### Create initial stacks

```{r}
#| label: create-stacks

# stack
enviro_vars_current <- stack(c(
  ocean_temp_present["thetao_mean_2"],
  #air_temp_present,
  ph_present["ph_mean_2"],
  #terrain,
  #rock_rast_combined,
  fetch_raster
))

names(enviro_vars_current) <- c(
  "ocean_temp",
  #"air_temp_2000",
  #"air_temp_2020",
  "ph",
  #"coastline_binary",
  #"rocky_substrate",
  "wave_fetch"
  )

names(enviro_vars_current)
plot(enviro_vars_current) # plots each decade

enviro_vars_future <- stack(c(
  ocean_temp_projected["thetao_mean_8"],
  #air_temp_projected,
  ph_projected["ph_mean_8"],
  #terrain,
  #rock_rast_combined,
  fetch_raster
))

names(enviro_vars_future) <- c(
  "ocean_temp",
  #"air_temp_2000",
  #"air_temp_2020",
  "ph",
  #"coastline_binary",
  #"rocky_substrate",
  "wave_fetch" # we need the future 
)

plot(enviro_vars_future) # plots each decade
```

### Check co-linearity

```{r}
#| label: check-colinearity

#pairs(sdmdata[,2:5], cex=0.1) #air temp and ocean temp highly correlate
#pairs(sdmdata[,5:8], cex=0.1) #pH with temp looks fine
```

We can see that air temp and ocean temp are highly correlated, hence we only want to use one of these variables. Ocean temp is a clear driver of biological coastal communities, so we've removed air temp.

## Creating SDM data sets

### Visualize occurrence over variables

Semibalanus

```{r}
#| label: view-obs-on-temp-semibalanus

# second layer of the SpatRaster (current ocean temp)
# we do not need the first column (species name)
final_visual_semibalanus  <- final_visual_semibalanus[,-1]

# plot
plot(enviro_vars_current, 3, col = cols)
lines(wrld)
points(final_visual_semibalanus$lon, final_visual_semibalanus$lat, col = 'blue')
```

Chthamalus

```{r}
#| label: view-obs-on-temp-chthamalus

# remove row
chthamalus_montagui_obs_both_methods <- chthamalus_montagui_obs_both_methods[,-1]

#plot
plot(enviro_vars_current, 1, col = cols)
lines(wrld)
points(chthamalus_montagui_obs_both_methods,
       col='blue')
```

### Fixed unmatched occurrence points to environmental data

We use the `extract` function from the `raster` package to check where they overlap.

```{r}
#| label: extract-values

# remove source column
final_visual_semibalanus_reduced <- final_visual_semibalanus[,-3]

#extract
presvals_semibalanus_visual <- extract(enviro_vars_current, final_visual_semibalanus_reduced) 

head(presvals_semibalanus_visual) # all NA!
```

We can see that some points are being treated as land rather than sea (we've got quite a few NAs). We need to find the closest point which is classed as coastline and use that instead.

Let's do Semibalanus first.

```{r}
#| label: correct_marclim_seminbalanus

# snap the valid environmental cells from function created earlier
semibalanus_corrected_marclim <- snap_to_nearest_valid(
  enviro_vars_current,
  semibalanus_marclim_broad)

# check
envir_terra <- rast(enviro_vars_current)
presvals_semibalanus_corrected_marclim <- extract(envir_terra, semibalanus_corrected_marclim) # this looks better

# remove the ID variable
presvals_semibalanus_corrected_marclim <- presvals_semibalanus_corrected_marclim[,-1]

# check the new data
head(semibalanus_corrected_marclim)

# for the PA version
marclim_broad_sp_PA_semibalanus <- marclim_broad_sp_PA %>% select(-c("Chthamalus.montagui"))

semibalanus_corrected_marclim_PA <- snap_to_nearest_valid(
  enviro_vars_current,
  marclim_broad_sp_PA_semibalanus)

# snap the valid environmental cells from function created earlier
presvals_semibalanus_corrected_marclim_PA <- extract(envir_terra, semibalanus_corrected_marclim_PA) # this looks better

# remove the ID variable
presvals_semibalanus_corrected_marclim_PA <- presvals_semibalanus_corrected_marclim_PA[,-1]

# check the new data
head(semibalanus_corrected_marclim_PA)

#add PA back in
semibalanus_corrected_marclim_PA$Semibalanus.balanoides = marclim_broad_sp_PA_semibalanus$Semibalanus.balanoides
```

```{r}
#| label: correct_gbif_seminbalanus

# snap the valid environmental cells from function created earlier
semibalanus_corrected_gbif <- snap_to_nearest_valid(
  enviro_vars_current,
  semibalanus_GBIF_clean_reduced)

# check
presvals_semibalanus_corrected_gbif <- extract(envir_terra, semibalanus_corrected_gbif[, c("lon", "lat")]) # this looks better

# remove the ID variable
presvals_semibalanus_corrected_gbif <- presvals_semibalanus_corrected_gbif[,-1]

# check the new data
head(semibalanus_corrected_gbif)

# Count how many presence points have all variables NA (or any NA)
na_points <- apply(presvals_semibalanus_corrected_gbif, 1, function(x) any(is.na(x)))

table(na_points) #no NAs
```

```{r}
#| label: correct_combined_seminbalanus

# snap the valid environmental cells from function created earlier
semibalanus_corrected_combined <- snap_to_nearest_valid(
  enviro_vars_current,
  semibalanus_GBIF_eDNA)

# check
presvals_semibalanus_corrected_combined <- extract(envir_terra, semibalanus_corrected_combined[, c("lon", "lat")]) # this looks better

# remove the ID variable
presvals_semibalanus_corrected_combined <- presvals_semibalanus_corrected_combined[,-1]

# check the new data
head(semibalanus_corrected_combined)

# Count how many presence points have all variables NA (or any NA)
na_points <- apply(presvals_semibalanus_corrected_combined, 1, function(x) any(is.na(x)))

table(na_points) #no NAs
```

Now let's do Chthamalus.

```{r}
#| label: correct_marclim_chthamalus

# snap the valid environmental cells from function created earlier
chthamalus_corrected_marclim <- snap_to_nearest_valid(
  enviro_vars_current,
  chthamalus_marclim_broad)

# check
envir_terra <- rast(enviro_vars_current)
presvals_chthamalus_corrected_marclim <- extract(envir_terra, chthamalus_corrected_marclim) # this looks better

# remove the ID variable
presvals_chthamalus_corrected_marclim <- presvals_chthamalus_corrected_marclim[,-1]

# check the new data
head(chthamalus_corrected_marclim)

# for the PA version

# call function we created eariler to correct points
marclim_broad_sp_PA_chthamalus <- marclim_broad_sp_PA %>% select(-c("Semibalanus.balanoides"))

# snap the valid environmental cells from function created earlier
chthamalus_corrected_marclim_PA <- snap_to_nearest_valid(
  enviro_vars_current,
  marclim_broad_sp_PA_chthamalus)

presvals_chthamalus_corrected_marclim_PA <- extract(envir_terra, chthamalus_corrected_marclim_PA) # this looks better

# remove the ID variable
presvals_chthamalus_corrected_marclim_PA <- presvals_chthamalus_corrected_marclim_PA[,-1]

# check the new data
head(chthamalus_corrected_marclim_PA)

#add PA back in
chthamalus_corrected_marclim_PA$Chthamalus.montagui = marclim_broad_sp_PA_chthamalus$Chthamalus.montagui
```

```{r}
#| label: correct_gbif_chthamalus

# snap the valid environmental cells from function created earlier
chthamalus_corrected_gbif <- snap_to_nearest_valid(
  enviro_vars_current,
  chthamalus_GBIF_clean_reduced)

# check
presvals_chthamalus_corrected_gbif <- extract(envir_terra, chthamalus_corrected_gbif[, c("lon", "lat")]) # this looks better

# remove the ID variable
presvals_chthamalus_corrected_gbif <- presvals_chthamalus_corrected_gbif[,-1]

# check the new data
head(chthamalus_corrected_gbif)

# Count how many presence points have all variables NA (or any NA)
na_points <- apply(presvals_chthamalus_corrected_gbif, 1, function(x) any(is.na(x)))

table(na_points) #no Nas
```

```{r}
#| label: correct_combined_chthamalus

# snap the valid environmental cells from function created earlier
chthamalus_corrected_combined <- snap_to_nearest_valid(
  enviro_vars_current,
  chthamalus_GBIF_eDNA)

# check
presvals_chthamalus_corrected_combined <- extract(envir_terra, chthamalus_corrected_combined[, c("lon", "lat")]) # this looks better

# remove the ID variable
presvals_chthamalus_corrected_combined <- presvals_chthamalus_corrected_combined[,-1]

# check the new data
head(chthamalus_corrected_combined)

# Count how many presence points have all variables NA (or any NA)
na_points <- apply(presvals_chthamalus_corrected_combined, 1, function(x) any(is.na(x)))

table(na_points) #no NAs
```

### Finalize environmental stacks after tidying

Let's now remove the coastline variables from the model.

```{r}
#| label: create-stacks-final

enviro_vars_current <- stack(c(
  ocean_temp_present["thetao_mean_2"],
  #air_temp_present,
  ph_present["ph_mean_2"],
  #terrain,
  #rock_rast_combined,
  fetch_raster
))

names(enviro_vars_current) <- c(
  "ocean_temp",
  #"air_temp_2000",
  #"air_temp_2020",
  "ph",
  #"coastline_binary",
  #"rocky_substrate",
  "wave_fetch")

plot(enviro_vars_current)
envir_terra <- rast(enviro_vars_current)

enviro_vars_future <- stack(c(
  ocean_temp_projected["thetao_mean_8"],
  #air_temp_projected,
  ph_projected["ph_mean_8"],
  #terrain,
  #rock_rast_combined,
  fetch_raster
))

names(enviro_vars_future) <- c(
  "ocean_temp",
  #"air_temp_2000",
  #"air_temp_2020",
  "ph",
  #"coastline_binary",
  #"rocky_substrate",
  "wave_fetch" # we need the future 
)

plot(enviro_vars_future) # plots each decade
```

## `biomod2` workflow

We are going to build our models and forecasts using the `biomod2` package. We will build four models, visualize their outputs and forecast changes into the far future (2100).

### Data formatting

For our models, the GBIF and eDNA data will be used to build the individual models, being split for example 10 times into 70/30 percentages to cross-validate them (which is defined in the next section).

The MarClim data will be used to evaluate all models (individual or ensemble) - this is defined when formatting the data in this section.

#### Data set 1 (Semibalanus - visual)

```{r}
#| label: semibalanus-visual-model-data-checks

# pa as df and tidy
# marclim
semibalanus_corrected_marclim_PA_df <- as.data.frame(semibalanus_corrected_marclim_PA)

# occurences
semibalanus_corrected_gbif_df <- as.data.frame(semibalanus_corrected_gbif)
semibalanus_corrected_gbif_df$Semibalanus.balanoides <- 1 #add p for gbif

# Extract env values
env_vals <- extract(envir_terra, semibalanus_corrected_gbif)

# Bind back to coordinates
pts_with_vals <- cbind(semibalanus_corrected_gbif_df, env_vals)

# Remove NAs
pts_clean <- pts_with_vals[complete.cases(pts_with_vals), c("lon", "lat")]
```

Now let's put all our data elements into the `BIOMOD_FormatingData` function.

```{r}
#| label: format-data-bodmod2-semibalanus-visual
#| echo: true
#| results: hide

myBiomodData_semibalanus_visual <- BIOMOD_FormatingData(
  resp.var = rep(1, nrow(pts_clean)),
  # presence-only
  expl.var = enviro_vars_current,
  resp.xy = pts_clean,
  resp.name = 'Semibalanus.balanoides',
  PA.nb.rep = 1,
  # Number of pseudo-absence sets (10 recom by biomd)
  PA.nb.absences = 3 * nrow(pts_clean),
  # Number of pseudo-absences (~3 x number of presences)
  PA.strategy = 'random',
  eval.resp.var = semibalanus_corrected_marclim_PA_df$Semibalanus.balanoides,
  # Evaluation responses (0/1)
  eval.resp.xy = semibalanus_corrected_marclim_PA_df[, c("lon", "lat")],
  # Evaluation locations,
  eval.expl.var = presvals_semibalanus_corrected_marclim_PA,
  #filter.raster = TRUE #autofilter,
  na.rm = FALSE
)

myBiomodData_semibalanus_visual

# Save BIOMOD_FormatingData plot to PNG
png("Figures/myBiomodData_semibalanus_visual.png", 
    width = 2000, height = 1500, res = 300)  # size in pixels, resolution in DPI
plot(myBiomodData_semibalanus_visual)
dev.off()
```

#### Data set 2 (Semibalanus - combined)

```{r}
#| label: semibalanus-combined-model-data-checks

# occurences
semibalanus_corrected_combined_df <- as.data.frame(semibalanus_corrected_combined)
semibalanus_corrected_combined_df$Semibalanus.balanoides <- 1 #add p for gbif

# Extract env values
env_vals <- extract(envir_terra, semibalanus_corrected_combined)

# Bind back to coordinates
pts_with_vals <- cbind(semibalanus_corrected_combined_df, env_vals)

# Remove NAs
pts_clean <- pts_with_vals[complete.cases(pts_with_vals), c("lon", "lat")]
```

Now let's put all our data elements into the `BIOMOD_FormatingData` function.

```{r}
#| label: format-data-bodmod2-semibalanus-combined
#| echo: true
#| results: hide

myBiomodData_semibalanus_combined <- BIOMOD_FormatingData(
  resp.var = rep(1, nrow(pts_clean)) ,
  # presence-only
  expl.var = enviro_vars_current,
  resp.xy = pts_clean,
  resp.name = 'Semibalanus.balanoides',
  PA.nb.rep = 1,
  # Number of pseudo-absence sets (10 recom by biomd)
  PA.nb.absences = 3 * nrow(pts_clean),
  # Number of pseudo-absences (~3 x number of presences)
  PA.strategy = 'random',
  eval.resp.var = semibalanus_corrected_marclim_PA_df$Semibalanus.balanoides,
  # Evaluation responses (0/1)
  eval.resp.xy = semibalanus_corrected_marclim_PA_df[, c("lon", "lat")],
  # Evaluation locations,
  eval.expl.var = presvals_semibalanus_corrected_marclim_PA
  #filter.raster = TRUE #autofilter
)

myBiomodData_semibalanus_combined

# Save BIOMOD_FormatingData plot to PNG
png("Figures/myBiomodData_semibalanus_combined.png", 
    width = 2000, height = 1500, res = 300)  # size in pixels, resolution in DPI
plot(myBiomodData_semibalanus_combined)
dev.off()
```

#### Data set 3 (Chthamalus - visual)

```{r}
#| label: chthamalus-visual-model-data-checks

# pa as df and tidy
# marclim
chthamalus_corrected_marclim_PA_df <- as.data.frame(chthamalus_corrected_marclim_PA)

# occurences
chthamalus_corrected_gbif_df <- as.data.frame(chthamalus_corrected_gbif)
chthamalus_corrected_gbif_df$Semibalanus.balanoides <- 1 #add p for gbif

# Extract env values
env_vals <- extract(envir_terra, chthamalus_corrected_gbif)

# Bind back to coordinates
pts_with_vals <- cbind(chthamalus_corrected_gbif_df, env_vals)

# Remove NAs
pts_clean <- pts_with_vals[complete.cases(pts_with_vals), c("lon", "lat")]
```

```{r}
#| label: format-data-bodmod2-chthamalus-visual
#| echo: true
#| results: hide

myBiomodData_chthamalus_visual <- BIOMOD_FormatingData(
  resp.var = rep(1, nrow(pts_clean)) ,
  # presence-only
  expl.var = enviro_vars_current,
  resp.xy = pts_clean,
  resp.name = 'Chthamalus.montagui',
  PA.nb.rep = 1,
  # Number of pseudo-absence sets (10 recom by biomd)
  PA.nb.absences = 3 * nrow(pts_clean),
  # Number of pseudo-absences (~3 x number of presences)
  PA.strategy = 'random',
  eval.resp.var = chthamalus_corrected_marclim_PA_df$Chthamalus.montagui,
  # Evaluation responses (0/1)
  eval.resp.xy = chthamalus_corrected_marclim_PA_df[, c("lon", "lat")],
  # Evaluation locations,
  eval.expl.var = presvals_chthamalus_corrected_marclim_PA
  #filter.raster = TRUE #autofilter
)

myBiomodData_chthamalus_visual

# Save BIOMOD_FormatingData plot to PNG
png("Figures/myBiomodData_chthamalus_visual.png", 
    width = 2000, height = 1500, res = 300)  # size in pixels, resolution in DPI
plot(myBiomodData_chthamalus_visual)
dev.off()
```

#### Data set 4 (Chthamalus - combined)

```{r}
#| label: chthamalus-combined-model-data-checks

# occurences
chthamalus_corrected_combined_df <- as.data.frame(chthamalus_corrected_combined)
chthamalus_corrected_combined_df$Semibalanus.balanoides <- 1 #add p for gbif

# Extract env values
env_vals <- extract(envir_terra, chthamalus_corrected_combined)

# Bind back to coordinates
pts_with_vals <- cbind(chthamalus_corrected_combined_df, env_vals)

# Remove NAs
pts_clean <- pts_with_vals[complete.cases(pts_with_vals), c("lon", "lat")]
```

```{r}
#| label: format-data-bodmod2-chthamalus-combined
#| echo: true
#| results: hide

myBiomodData_chthamalus_combined <- BIOMOD_FormatingData(
  resp.var = rep(1, nrow(pts_clean)) ,
  # presence-only
  expl.var = enviro_vars_current,
  resp.xy = pts_clean,
  resp.name = 'Chthamalus.montagui',
  PA.nb.rep = 1,
  # Number of pseudo-absence sets (10 recom by biomd)
  PA.nb.absences = 3 * nrow(pts_clean),
  # Number of pseudo-absences (~3 x number of presences)
  PA.strategy = 'random',
  eval.resp.var = chthamalus_corrected_marclim_PA_df$Chthamalus.montagui,
  # Evaluation responses (0/1)
  eval.resp.xy = chthamalus_corrected_marclim_PA_df[, c("lon", "lat")],
  # Evaluation locations,
  eval.expl.var = presvals_chthamalus_corrected_marclim_PA
  #filter.raster = TRUE #autofilter
)

myBiomodData_chthamalus_combined

# Save BIOMOD_FormatingData plot to PNG
png("Figures/myBiomodData_chthamalus_combined.png", 
    width = 2000, height = 1500, res = 300)  # size in pixels, resolution in DPI
plot(myBiomodData_chthamalus_combined)
dev.off()
```

We now have four `biomod` objects ready for modelling.

```{r}
#| label: biomod-objects

summary(myBiomodData_semibalanus_visual)
summary(myBiomodData_semibalanus_combined)
summary(myBiomodData_chthamalus_visual)
summary(myBiomodData_chthamalus_combined)
```

### Run models

Time to run our models!

#### Model 1 (Semibalanus - visual)

```{r}
#| label: run-model-semibalanus-visual-single
#| echo: true
#| results: hide

# Model single models
myBiomodModelOut_semibalanus_visual <- BIOMOD_Modeling(
  bm.format = myBiomodData_semibalanus_visual,
  models = c('GAM','GLM','RF','MAXENT'),
  modeling.id = "Semibalanus.visual",
  CV.strategy = 'random', #cross-validation selection strategy
  CV.nb.rep = 2, #number of sets of cross-validation points
  CV.perc = 0.7, #percentage kept for calibration
  OPT.strategy = 'default',
  var.import = 2,
  metric.eval = c('TSS', 'ROC')
)

# check which options can be altered
get_options(myBiomodModelOut_semibalanus_visual)

# see output
myBiomodModelOut_semibalanus_visual
```

```{r}
#| label: explore-model-semibalanus-visual-single
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
head(get_evaluations(myBiomodModelOut_semibalanus_visual))
head(get_variables_importance(myBiomodModelOut_semibalanus_visual))

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut_semibalanus_visual)
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_semibalanus_visual, group.by = c('algo', 'algo'))
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_semibalanus_visual, group.by = c('algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_semibalanus_visual, group.by = c('expl.var', 'algo', 'algo'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_semibalanus_visual, group.by = c('expl.var', 'algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_semibalanus_visual, group.by = c('algo', 'expl.var', 'run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut_semibalanus_visual, 
                      models.chosen = get_built_models(myBiomodModelOut_semibalanus_visual)[c(1:3, 12:14)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_semibalanus_visual, 
                      models.chosen = get_built_models(myBiomodModelOut_semibalanus_visual)[c(1:3, 12:14)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_semibalanus_visual, 
                      models.chosen = get_built_models(myBiomodModelOut_semibalanus_visual)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

```{r}
#| label: run-model-semibalanus-visual-ensemble
#| echo: true
#| results: hide

# Model ensemble models
myBiomodEM_semibalanus_visual <- BIOMOD_EnsembleModeling(
  bm.mod = myBiomodModelOut_semibalanus_visual,
  models.chosen = 'all',
  em.by = 'all',
  em.algo = c('EMmean', 'EMcv', 'EMci', 'EMca', 'EMwmean'),
  metric.select = c('TSS', 'ROC'),
  metric.select.thresh = c(0.4, 0.7),
  metric.eval = c('TSS', 'ROC'),
  var.import = 2,
  EMci.alpha = 0.05,
  EMwmean.decay = 'proportional'
)

myBiomodEM_semibalanus_visual
```

```{r}
#| label: explore-model-semibalanus-visual-ensemble
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodEM_semibalanus_visual)
get_variables_importance(myBiomodEM_semibalanus_visual)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodEM_semibalanus_visual, group.by = 'full.name')
bm_PlotEvalBoxplot(bm.out = myBiomodEM_semibalanus_visual, group.by = c('full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_semibalanus_visual, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_semibalanus_visual, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_semibalanus_visual, group.by = c('algo', 'expl.var', 'merged.by.run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM_semibalanus_visual, 
                      models.chosen = get_built_models(myBiomodEM_semibalanus_visual)[c(1, 6, 7)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodEM_semibalanus_visual, 
                      models.chosen = get_built_models(myBiomodEM_semibalanus_visual)[c(1, 6, 7)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodEM_semibalanus_visual, 
                      models.chosen = get_built_models(myBiomodEM_semibalanus_visual)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

#### Model 2 (Semibalanus - combined)

```{r}
#| label: run-model-semibalanus-combined-single
#| echo: true
#| results: hide

# Model single models
myBiomodModelOut_semibalanus_combined <- BIOMOD_Modeling(
  bm.format = myBiomodData_semibalanus_combined,
  models = c('GAM','GLM','RF','MAXENT'),
  modeling.id = "Semibalanus.combined",
  CV.strategy = 'random',
  CV.nb.rep = 2,
  CV.perc = 0.7,
  OPT.strategy = 'default',
  var.import = 2,
  metric.eval = c('TSS', 'ROC')
)

get_options(myBiomodModelOut_semibalanus_combined)

myBiomodModelOut_semibalanus_combined
```

```{r}
#| label: explore-model-semibalanus-combined-single
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodModelOut_semibalanus_combined)
get_variables_importance(myBiomodModelOut_semibalanus_combined)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut_semibalanus_combined)
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_semibalanus_combined, group.by = c('algo', 'algo'))
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_semibalanus_combined, group.by = c('algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_semibalanus_combined, group.by = c('expl.var', 'algo', 'algo'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_semibalanus_combined, group.by = c('expl.var', 'algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_semibalanus_combined, group.by = c('algo', 'expl.var', 'run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut_semibalanus_combined, 
                      models.chosen = get_built_models(myBiomodModelOut_semibalanus_combined)[c(1:3, 12:14)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_semibalanus_combined, 
                      models.chosen = get_built_models(myBiomodModelOut_semibalanus_combined)[c(1:3, 12:14)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_semibalanus_combined, 
                      models.chosen = get_built_models(myBiomodModelOut_semibalanus_combined)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

```{r}
#| label: run-model-semibalanus-combined-ensemble
#| echo: true
#| results: hide

# Model ensemble models
myBiomodEM_semibalanus_combined <- BIOMOD_EnsembleModeling(
  bm.mod = myBiomodModelOut_semibalanus_combined,
  models.chosen = 'all',
  em.by = 'all',
  em.algo = c('EMmean', 'EMcv', 'EMci', 'EMca', 'EMwmean'),
  metric.select = c('TSS', 'ROC'),
  metric.select.thresh = c(0.3, 0.6),
  metric.eval = c('TSS', 'ROC'),
  var.import = 2,
  EMci.alpha = 0.05,
  EMwmean.decay = 'proportional'
)

myBiomodEM_semibalanus_combined
```

```{r}
#| label: explore-model-semibalanus-combined-ensemble
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodEM_semibalanus_combined)
get_variables_importance(myBiomodEM_semibalanus_combined)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodEM_semibalanus_combined, group.by = 'full.name')
bm_PlotEvalBoxplot(bm.out = myBiomodEM_semibalanus_combined, group.by = c('full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_semibalanus_combined, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_semibalanus_combined, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_semibalanus_combined, group.by = c('algo', 'expl.var', 'merged.by.run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM_semibalanus_combined, 
                      models.chosen = get_built_models(myBiomodEM_semibalanus_combined)[c(1, 6, 7)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodEM_semibalanus_combined, 
                      models.chosen = get_built_models(myBiomodEM_semibalanus_combined)[c(1, 6, 7)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodEM_semibalanus_combined, 
                      models.chosen = get_built_models(myBiomodEM_semibalanus_combined)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

#### Model 3 (Chthamalus - visual)

```{r}
#| label: run-model-chthamalus-visual-single
#| echo: true
#| results: hide

# Model single models
myBiomodModelOut_chthamalus_visual <- BIOMOD_Modeling(
  bm.format = myBiomodData_chthamalus_visual,
  models = c('GAM','GLM','RF','MAXENT'),
  modeling.id = "chthamalus.visual",
  CV.strategy = 'random',
  CV.nb.rep = 2,
  CV.perc = 0.7,
  OPT.strategy = 'default',
  var.import = 2,
  metric.eval = c('TSS', 'ROC')
)

get_options(myBiomodModelOut_chthamalus_visual)

myBiomodModelOut_chthamalus_visual
```

```{r}
#| label: explore-model-chthamalus-visual-single
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodModelOut_chthamalus_visual)
get_variables_importance(myBiomodModelOut_chthamalus_visual)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut_chthamalus_visual)
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_chthamalus_visual, group.by = c('algo', 'algo'))
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_chthamalus_visual, group.by = c('algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_chthamalus_visual, group.by = c('expl.var', 'algo', 'algo'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_chthamalus_visual, group.by = c('expl.var', 'algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_chthamalus_visual, group.by = c('algo', 'expl.var', 'run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut_chthamalus_visual, 
                      models.chosen = get_built_models(myBiomodModelOut_chthamalus_visual)[c(1:3, 12:14)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_chthamalus_visual, 
                      models.chosen = get_built_models(myBiomodModelOut_chthamalus_visual)[c(1:3, 12:14)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_chthamalus_visual, 
                      models.chosen = get_built_models(myBiomodModelOut_chthamalus_visual)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

```{r}
#| label: run-model-chthamalus-visual-ensemble
#| echo: true
#| results: hide

# Model ensemble models (takes around 30 mins)
myBiomodEM_chthamalus_visual <- BIOMOD_EnsembleModeling(
  bm.mod = myBiomodModelOut_chthamalus_visual,
  models.chosen = 'all',
  em.by = 'all',
  em.algo = c('EMmean', 'EMcv', 'EMci', 'EMca', 'EMwmean'),
  metric.select = c('TSS', 'ROC'),
  metric.select.thresh = c(0.3, 0.6),
  metric.eval = c('TSS', 'ROC'),
  var.import = 2,
  EMci.alpha = 0.05,
  EMwmean.decay = 'proportional'
)

myBiomodEM_chthamalus_visual
```

```{r}
#| label: explore-model-chthamalus-visual-ensemble
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodEM_chthamalus_visual)
get_variables_importance(myBiomodEM_chthamalus_visual)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodEM_chthamalus_visual, group.by = 'full.name')
bm_PlotEvalBoxplot(bm.out = myBiomodEM_chthamalus_visual, group.by = c('full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_chthamalus_visual, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_chthamalus_visual, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_chthamalus_visual, group.by = c('algo', 'expl.var', 'merged.by.run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM_chthamalus_visual, 
                      models.chosen = get_built_models(myBiomodEM_chthamalus_visual)[c(1, 6, 7)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodEM_chthamalus_visual, 
                      models.chosen = get_built_models(myBiomodEM_chthamalus_visual)[c(1, 6, 7)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodEM_chthamalus_visual, 
                      models.chosen = get_built_models(myBiomodEM_chthamalus_visual)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

#### Model 4 (Chthamalus - combined)

```{r}
#| label: run-model-chthamalus-combined-single
#| echo: true
#| results: hide

# Model single models
myBiomodModelOut_chthamalus_combined <- BIOMOD_Modeling(
  bm.format = myBiomodData_chthamalus_combined,
  models = c('GAM','GLM','RF','MAXENT'),
  modeling.id = "chthamalus.combined",
  CV.strategy = 'random',
  CV.nb.rep = 2,
  CV.perc = 0.7,
  OPT.strategy = 'default',
  var.import = 2,
  metric.eval = c('TSS', 'ROC')
)

get_options(myBiomodModelOut_chthamalus_combined)

myBiomodModelOut_chthamalus_combined
```

```{r}
#| label: explore-model-chthamalus-combined-single
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodModelOut_chthamalus_combined)
get_variables_importance(myBiomodModelOut_chthamalus_combined)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut_chthamalus_combined)
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_chthamalus_combined, group.by = c('algo', 'algo'))
bm_PlotEvalBoxplot(bm.out = myBiomodModelOut_chthamalus_combined, group.by = c('algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_chthamalus_combined, group.by = c('expl.var', 'algo', 'algo'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_chthamalus_combined, group.by = c('expl.var', 'algo', 'run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut_chthamalus_combined, group.by = c('algo', 'expl.var', 'run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut_chthamalus_combined, 
                      models.chosen = get_built_models(myBiomodModelOut_chthamalus_combined)[c(1:3, 12:14)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_chthamalus_combined, 
                      models.chosen = get_built_models(myBiomodModelOut_chthamalus_combined)[c(1:3, 12:14)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodModelOut_chthamalus_combined, 
                      models.chosen = get_built_models(myBiomodModelOut_chthamalus_combined)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

```{r}
#| label: run-model-chthamalus-combined-ensemble
#| echo: true
#| results: hide

# Model ensemble models (takes around 30 mins)
myBiomodEM_chthamalus_combined <- BIOMOD_EnsembleModeling(
  bm.mod = myBiomodModelOut_chthamalus_combined,
  models.chosen = 'all',
  em.by = 'all',
  em.algo = c('EMmean', 'EMcv', 'EMci', 'EMca', 'EMwmean'),
  metric.select = c('TSS', 'ROC'),
  metric.select.thresh = c(0.3, 0.6),
  metric.eval = c('TSS', 'ROC'),
  var.import = 2,
  EMci.alpha = 0.05,
  EMwmean.decay = 'proportional'
)

myBiomodEM_chthamalus_combined
```

```{r}
#| label: explore-model-chthamalus-combined-ensemble
#| echo: true
#| results: hide

# Get evaluation scores & variables importance
get_evaluations(myBiomodEM_chthamalus_combined)
get_variables_importance(myBiomodEM_chthamalus_combined)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodEM_chthamalus_combined, group.by = 'full.name')
bm_PlotEvalBoxplot(bm.out = myBiomodEM_chthamalus_combined, group.by = c('full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_chthamalus_combined, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_chthamalus_combined, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM_chthamalus_combined, group.by = c('algo', 'expl.var', 'merged.by.run'))

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM_chthamalus_combined, 
                      models.chosen = get_built_models(myBiomodEM_chthamalus_combined)[c(1, 6, 7)],
                      fixed.var = 'median')
bm_PlotResponseCurves(bm.out = myBiomodEM_chthamalus_combined, 
                      models.chosen = get_built_models(myBiomodEM_chthamalus_combined)[c(1, 6, 7)],
                      fixed.var = 'min')
bm_PlotResponseCurves(bm.out = myBiomodEM_chthamalus_combined, 
                      models.chosen = get_built_models(myBiomodEM_chthamalus_combined)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

### Current projections

We are going to run projections and mask our rocky substrate layer over projections. This ensures that any areas without rocky habitats won't be included in our final maps.

#### Projection 1 (Semibalanus - visual)

```{r}
#| label: predict-model-semibalanus-visual
#| echo: true
#| results: hide

#just using mean models from now on

myBiomodEMProj_semibalanus_visual <- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_semibalanus_visual,
  proj.name = 'CurrentEM.semibalanus.visual',
  new.env = enviro_vars_current,
  models.chosen = c(
    'Semibalanus.balanoides_EMmeanByTSS_mergedData_mergedRun_mergedAlgo', 
    'Semibalanus.balanoides_EMmeanByROC_mergedData_mergedRun_mergedAlgo')
  #metric.binary = 'all',
  #metric.filter = 'all'
)

myBiomodEMProj_semibalanus_visual

# visualise
proj_semibalanus_visual <- get_predictions(myBiomodEMProj_semibalanus_visual) # returns a rasterLayer
plot(proj_semibalanus_visual)

# after rocky mask
proj_semibalanus_visual_masked <- mask(proj_semibalanus_visual, expanded_rock)
plot(proj_semibalanus_visual_masked)
```

```{r}
#| label: plots-current-protection-semibal-visual

# Choose the first layer
r_layer <- proj_semibalanus_visual_masked[[1]] 

# Convert SpatRaster to data frame for ggplot
r_df <- as.data.frame(r_layer, xy = TRUE)
colnames(r_df) <- c("x", "y", "suitability")

# Load UK outline
uk <- rnaturalearth::ne_countries(country = "United Kingdom",
                                  scale = "large",
                                  returnclass = "sf")
# plot
semibalanus_visual_current_plot_TSS <- ggplot() +
  geom_sf(
    data = uk,
    fill = "grey98",       # land
    color = "grey80",
    size = 0.7
  ) +
  geom_tile(data = r_df, aes(x = x, y = y, fill = suitability)) +
  scale_fill_viridis_c(option = "viridis", na.value = "transparent") +
  coord_sf(
    xlim = c(-11, 2.6),
    ylim = c(49.35, 60.9),
    expand = FALSE
  ) +
  theme_minimal(base_size = 14) +
  labs(fill = NULL) +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "npc"),   # ~80% of plotting height
      barwidth  = unit(0.02, "npc"),  # slim vertical bar
      title.position = "top"
    )
  ) +
  theme(
    legend.position = "right",        # place bar to the right
    axis.title.x = element_blank(),   # remove x-axis title
    axis.title.y = element_blank(),   # remove y-axis title
    panel.background = element_rect(fill = "white", colour = NA),  
    plot.background = element_rect(fill = "white", colour = NA)     
  )

semibalanus_visual_current_plot_TSS
```

#### Projection 2 (Semibalanus - combined)

```{r}
#| label: predict-model-semibalanus-combined
#| echo: true
#| results: hide

myBiomodEMProj_semibalanus_combined <- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_semibalanus_combined,
  proj.name = 'CurrentEM.semibalanus.combined',
  new.env = enviro_vars_current,
  models.chosen = c(
    'Semibalanus.balanoides_EMmeanByTSS_mergedData_mergedRun_mergedAlgo', 
    'Semibalanus.balanoides_EMmeanByROC_mergedData_mergedRun_mergedAlgo')
  #metric.binary = 'all',
  #metric.filter = 'all'
)

myBiomodEMProj_semibalanus_combined

# visualise
proj_semibalanus_combined <- get_predictions(myBiomodEMProj_semibalanus_combined) # returns a rasterLayer
plot(proj_semibalanus_combined)

# after rocky mask
proj_semibalanus_combined_masked <- mask(proj_semibalanus_combined, expanded_rock)
plot(proj_semibalanus_combined_masked)
```

```{r}
#| label: plots-current-protection-semibal-combined

# Choose the first layer
r_layer <- proj_semibalanus_combined_masked[[1]] 

# Convert SpatRaster to data frame for ggplot
r_df <- as.data.frame(r_layer, xy = TRUE)
colnames(r_df) <- c("x", "y", "suitability")

# plot
semibalanus_combined_current_plot_TSS <- ggplot() +
  geom_sf(
    data = uk,
    fill = "grey98",       # land
    color = "grey80",
    size = 0.7
  ) +
  geom_tile(data = r_df, aes(x = x, y = y, fill = suitability)) +
  scale_fill_viridis_c(option = "viridis", na.value = "transparent") +
  labs(fill = NULL) + 
  coord_sf(
    xlim = c(-11, 2.6),
    ylim = c(49.35, 60.9),
    expand = FALSE
  ) +
  theme_minimal(base_size = 14) +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "npc"),   # ~80% of plotting height
      barwidth  = unit(0.02, "npc")   # slim vertical bar
    )
  ) +
  theme(
    legend.position = "right",        # put the colorbar on the right
    axis.title.x = element_blank(),   # remove x-axis title
    axis.title.y = element_blank(),   # remove y-axis title
    panel.background = element_rect(fill = "white", colour = NA),  
    plot.background = element_rect(fill = "white", colour = NA)     
  )

semibalanus_combined_current_plot_TSS
```

#### Projection 3 (Chthamalus - visual)

```{r}
#| label: predict-model-chthamalus-visual
#| echo: true
#| results: hide

myBiomodEMProj_chthamalus_visual <- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_chthamalus_visual,
  proj.name = 'CurrentEM.chthamalus.visual',
  new.env = enviro_vars_current,
  models.chosen = c(
    'Chthamalus.montagui_EMmeanByTSS_mergedData_mergedRun_mergedAlgo', 
    'Chthamalus.montagui_EMmeanByROC_mergedData_mergedRun_mergedAlgo')
  #metric.binary = 'all',
  #metric.filter = 'all'
)

myBiomodEMProj_chthamalus_visual

# visualise
proj_chthamalus_visual <- get_predictions(myBiomodEMProj_chthamalus_visual) # returns a rasterLayer
plot(proj_chthamalus_visual)

# after rocky mask
proj_chthamalus_visual_masked <- mask(proj_chthamalus_visual, expanded_rock)
plot(proj_chthamalus_visual_masked)
```

```{r}
#| label: plots-current-protection-chthamalus-visual

# Choose the first layer
r_layer <- proj_chthamalus_visual_masked[[1]] 

# Convert SpatRaster to data frame for ggplot
r_df <- as.data.frame(r_layer, xy = TRUE)
colnames(r_df) <- c("x", "y", "suitability")

# plot
chthamalus_visual_current_plot_TSS <- ggplot() +
  geom_sf(
    data = uk,
    fill = "grey98",       # land
    color = "grey80",
    size = 0.7
  ) +
  geom_tile(data = r_df, aes(x = x, y = y, fill = suitability)) +
  scale_fill_viridis_c(option = "viridis", na.value = "transparent") +
  labs(fill = NULL) + 
  coord_sf(
    xlim = c(-11, 2.6),
    ylim = c(49.35, 60.9),
    expand = FALSE
  ) +
  theme_minimal(base_size = 14) +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "npc"),   # ~80% of plotting height
      barwidth  = unit(0.02, "npc")   # slim vertical bar
    )
  ) +
  theme(
    legend.position = "right",        # show the colorbar on the right
    axis.title.x = element_blank(),   # remove x-axis title
    axis.title.y = element_blank(),   # remove y-axis title
    panel.background = element_rect(fill = "white", colour = NA),  
    plot.background = element_rect(fill = "white", colour = NA)     
  )

chthamalus_visual_current_plot_TSS
```

#### Projection 4 (Chthamalus - combined)

```{r}
#| label: predict-model-chthamalus-combined
#| echo: true
#| results: hide

myBiomodEMProj_chthamalus_combined <- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_chthamalus_combined,
  proj.name = 'CurrentEM.chthamalus.combined',
  new.env = enviro_vars_current,
  models.chosen = c(
    'Chthamalus.montagui_EMmeanByTSS_mergedData_mergedRun_mergedAlgo', 
    'Chthamalus.montagui_EMmeanByROC_mergedData_mergedRun_mergedAlgo')
  #metric.binary = 'all',
  #metric.filter = 'all'
)

myBiomodEMProj_chthamalus_combined

# visualise
proj_chthamalus_combined <- get_predictions(myBiomodEMProj_chthamalus_combined) # returns a rasterLayer
plot(proj_chthamalus_combined)

# after rocky mask
proj_chthamalus_combined_masked <- mask(proj_chthamalus_combined, expanded_rock)
plot(proj_chthamalus_combined_masked)
```

```{r}
#| label: plots-current-protection-chthamalus-combined

# Choose the first layer
r_layer <- proj_chthamalus_combined_masked[[1]] 

# Convert SpatRaster to data frame for ggplot
r_df <- as.data.frame(r_layer, xy = TRUE)
colnames(r_df) <- c("x", "y", "suitability")

#plot
chthamalus_combined_current_plot_TSS <- ggplot() +
  geom_sf(
    data = uk,
    fill = "grey98",       # land
    color = "grey80",
    size = 0.7
  ) +
  geom_tile(data = r_df, aes(x = x, y = y, fill = suitability)) +
  scale_fill_viridis_c(option = "viridis", na.value = "transparent") +
  labs(fill = NULL) +   # 🚨 remove legend title
  coord_sf(
    xlim = c(-11, 2.6),
    ylim = c(49.35, 60.9),
    expand = FALSE
  ) +
  theme_minimal(base_size = 14) +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.5, "npc"),   # ~80% of plot height
      barwidth  = unit(0.02, "npc")   # slim vertical bar
    )
  ) +
  theme(
    legend.position = "right",        # show colorbar on the right
    axis.title.x = element_blank(),   # remove x-axis title
    axis.title.y = element_blank(),   # remove y-axis title
    panel.background = element_rect(fill = "white", colour = NA),  
    plot.background = element_rect(fill = "white", colour = NA)     
  )

chthamalus_combined_current_plot_TSS
```

#### Joint plot

Let's plot all our current projections together into one plot.

```{r}
#| label: current-protections-plot

current_projections_plot <- cowplot::plot_grid(
  semibalanus_visual_current_plot_TSS,
  semibalanus_combined_current_plot_TSS,
  chthamalus_visual_current_plot_TSS,
  chthamalus_combined_current_plot_TSS, ncol = 2, labels = c("a) SB-Citizen", "b) SB-Combined", "c) CM-Citizen", "d) CM-Combined"), align = "hv"
)

current_projections_plot

# Save the plot
ggsave(current_projections_plot, filename = "Figures/current_projections_plot.png", width = 8, height = 9, dpi = 300)
```

### Forecasting change

#### Forecast 1 (Semibalanus - visual)

```{r}
#| label: forecast-model-semibalanus-visual-EM

myBiomodEMProjFuture_semibalanus_visual <- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_semibalanus_visual,
  proj.name = 'FutureEM.semibalanus.visual',
  new.env = enviro_vars_future,
  models.chosen = c(
    'Semibalanus.balanoides_EMmeanByTSS_mergedData_mergedRun_mergedAlgo',
    "Semibalanus.balanoides_EMmeanByROC_mergedData_mergedRun_mergedAlgo"
  ),
  metric.binary = 'all',
  metric.filter = 'all'
)

myBiomodEMProjFuture_semibalanus_visual
plot(myBiomodEMProjFuture_semibalanus_visual)
```

```{r}
#| label: compare-range-sizes_semibalanus-visual

# Project onto future conditions
myBiomodProjectionFuture_semibalanus_visual <- BIOMOD_Projection(
  bm.mod = myBiomodModelOut_semibalanus_visual,
  proj.name = 'Future.single.semibalanus.visual',
  new.env = enviro_vars_future,
  models.chosen = 'all',
  metric.binary = 'TSS',
  build.clamping.mask = TRUE
)

# Load current and future binary projections
CurrentProj_semibalanus_visual <- get_predictions(myBiomodProjectionFuture_semibalanus_visual, metric.binary = "TSS")
FutureProj_semibalanus_visual <- get_predictions(myBiomodProjectionFuture_semibalanus_visual, metric.binary = "TSS")

# Compute differences
myBiomodRangeSize_semibalanus_visual <- BIOMOD_RangeSize(proj.current = CurrentProj_semibalanus_visual, proj.future = FutureProj_semibalanus_visual)

myBiomodRangeSize_semibalanus_visual$Compt.By.Models
plot(myBiomodRangeSize_semibalanus_visual$Diff.By.Pixel)

# Represent main results
gg_semibalanus_visual = bm_PlotRangeSize(
  bm.range = myBiomodRangeSize_semibalanus_visual,
  do.count = TRUE,
  do.perc = TRUE,
  do.maps = TRUE,
  do.mean = TRUE,
  do.plot = TRUE,
  row.names = c("Species", "Dataset", "Run", "Algo")
)
```

#### Forecast 2 (Semibalanus - combined)

```{r}
#| label: forecast-model-semibalanus-combined

myBiomodEMProjFuture_semibalanus_combined <- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_semibalanus_combined,
  proj.name = 'FutureEM.semibalanus.combined',
  new.env = enviro_vars_future,
  models.chosen = c(
    'Semibalanus.balanoides_EMmeanByTSS_mergedData_mergedRun_mergedAlgo',
    "Semibalanus.balanoides_EMmeanByROC_mergedData_mergedRun_mergedAlgo"
  ),
  metric.binary = 'all',
  metric.filter = 'all'
)

myBiomodEMProjFuture_semibalanus_combined
plot(myBiomodEMProjFuture_semibalanus_combined)
```

```{r}
#| label: compare-range-sizes_semibalanus-combined

# Project onto future conditions
myBiomodProjectionFuture_semibalanus_combined <- BIOMOD_Projection(
  bm.mod = myBiomodModelOut_semibalanus_combined,
  proj.name = 'Future.single.semibalanus.combined',
  new.env = enviro_vars_future,
  models.chosen = 'all',
  metric.binary = 'TSS',
  build.clamping.mask = TRUE
)

# Load current and future binary projections
CurrentProj_semibalanus_combined <- get_predictions(myBiomodProjectionFuture_semibalanus_combined,
                                                    metric.binary = "TSS")
FutureProj_semibalanus_combined <- get_predictions(myBiomodProjectionFuture_semibalanus_combined,
                                                   metric.binary = "TSS")

# Compute differences
myBiomodRangeSize_semibalanus_combined <- BIOMOD_RangeSize(proj.current = CurrentProj_semibalanus_combined, proj.future = FutureProj_semibalanus_combined)

myBiomodRangeSize_semibalanus_combined$Compt.By.Models
plot(myBiomodRangeSize_semibalanus_combined$Diff.By.Pixel)

# Represent main results
gg_semibalanus_combined = bm_PlotRangeSize(
  bm.range = myBiomodRangeSize_semibalanus_combined,
  do.count = TRUE,
  do.perc = TRUE,
  do.maps = TRUE,
  do.mean = TRUE,
  do.plot = TRUE,
  row.names = c("Species", "Dataset", "Run", "Algo")
)
```

#### Forecast 3 (Chthamalus - visual)

```{r}
#| label: forecast-model-chthamalus-visual

myBiomodEMProjFuture_chthamalus_visual<- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_chthamalus_visual,
  proj.name = 'FutureEM.chthamalus.visual',
  new.env = enviro_vars_future,
  models.chosen = c(
    'Chthamalus.montagui_EMmeanByTSS_mergedData_mergedRun_mergedAlgo',
    "Chthamalus.montagui_EMmeanByROC_mergedData_mergedRun_mergedAlgo"
  ),
  metric.binary = 'all',
  metric.filter = 'all'
)

myBiomodEMProjFuture_chthamalus_visual
plot(myBiomodEMProjFuture_chthamalus_visual)
```

```{r}
#| label: compare-range-sizes_chthamalus-visual

# Project onto future conditions
myBiomodProjectionFuture_chthamalus_visual <- BIOMOD_Projection(
  bm.mod = myBiomodModelOut_chthamalus_visual,
  proj.name = 'Future.single.chthamalus.visual',
  new.env = enviro_vars_future,
  models.chosen = 'all',
  metric.binary = 'TSS',
  build.clamping.mask = TRUE
)

# Load current and future binary projections
CurrentProj_chthamalus_visual <- get_predictions(myBiomodProjectionFuture_chthamalus_visual, metric.binary = "TSS")
FutureProj_chthamalus_visual <- get_predictions(myBiomodProjectionFuture_chthamalus_visual, metric.binary = "TSS")

# Compute differences
myBiomodRangeSize_chthamalus_visual <- BIOMOD_RangeSize(proj.current = CurrentProj_chthamalus_visual, proj.future = FutureProj_chthamalus_visual)

myBiomodRangeSize_chthamalus_visual$Compt.By.Models
plot(myBiomodRangeSize_chthamalus_visual$Diff.By.Pixel)

# Represent main results
gg_chthamalus_visual = bm_PlotRangeSize(
  bm.range = myBiomodRangeSize_chthamalus_visual,
  do.count = TRUE,
  do.perc = TRUE,
  do.maps = TRUE,
  do.mean = TRUE,
  do.plot = TRUE,
  row.names = c("Species", "Dataset", "Run", "Algo")
)
```

#### Forecast 4 (Chthamalus - combined)

```{r}
#| label: forecast-model-chthamalus-combined

myBiomodEMProjFuture_chthamalus_combined<- BIOMOD_EnsembleForecasting(
  bm.em = myBiomodEM_chthamalus_combined,
  proj.name = 'FutureEM.chthamalus.combined',
  new.env = enviro_vars_future,
  models.chosen = c(
    'Chthamalus.montagui_EMmeanByTSS_mergedData_mergedRun_mergedAlgo',
    "Chthamalus.montagui_EMmeanByROC_mergedData_mergedRun_mergedAlgo"
  ),
  metric.binary = 'all',
  metric.filter = 'all'
)

myBiomodEMProjFuture_chthamalus_combined
plot(myBiomodEMProjFuture_chthamalus_combined)
```

```{r}
#| label: compare-range-sizes_chthamalus-combined

# Project onto future conditions
myBiomodProjectionFuture_chthamalus_combined <- BIOMOD_Projection(
  bm.mod = myBiomodModelOut_chthamalus_combined,
  proj.name = 'Future.single.chthamalus.combined',
  new.env = enviro_vars_future,
  models.chosen = 'all',
  metric.binary = 'TSS',
  build.clamping.mask = TRUE
)

# Load current and future binary projections
CurrentProj_chthamalus_combined <- get_predictions(myBiomodProjectionFuture_chthamalus_combined, metric.binary = "TSS")
FutureProj_chthamalus_combined <- get_predictions(myBiomodProjectionFuture_chthamalus_combined, metric.binary = "TSS")

# Compute differences
myBiomodRangeSize_chthamalus_combined <- BIOMOD_RangeSize(proj.current = CurrentProj_chthamalus_combined, proj.future = FutureProj_chthamalus_combined)

myBiomodRangeSize_chthamalus_combined$Compt.By.Models
plot(myBiomodRangeSize_chthamalus_combined$Diff.By.Pixel)

# Represent main results
gg_chthamalus_combined = bm_PlotRangeSize(
  bm.range = myBiomodRangeSize_chthamalus_combined,
  do.count = TRUE,
  do.perc = TRUE,
  do.maps = TRUE,
  do.mean = TRUE,
  do.plot = TRUE,
  row.names = c("Species", "Dataset", "Run", "Algo")
)
```
